Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-24 write-ups/FSE/paper.tex
Shortening with Tyler.

\newcommand{\myparagraph}[1]{\paragraph{#1}}
%{\noindent{\bf\emph{#1}}\hspace{1em}}

@@

\myparagraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating
such algorithms on GPUs, particularly the ones that
traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  For example, many graph algorithms employ a level-by-level
strategy, requiring a global barrier between each level, and work
stealing algorithms require each workgroup to maintain a queue, which
is typically protected by a mutex to enable stealing by other
workgroups.

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.

%\vspace{-1mm}
\myparagraph{A degree of fairness: occupancy-bound execution} GPU
schedulers are \emph{not} fair. The current GPU programming
models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding
scheduling of workgroups, and current GPU scheduler are unfair in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.\footnote{In practice,
  depending on the kernel, multiple workgroups might map to the same
  compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle more workgroups being
launched than there are compute units is via an \emph{occupancy-bound}
execution
model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where,
once a workgroup has commenced execution on a compute unit (it has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it finishes execution.  Experiments suggest that
the occupancy-bound execution model is widely implemented by today's
GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.
It is evidently a simple model for device vendors to implement
efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, which are
bound to separate compute units that operate in parallel.  Current GPU
implementations of blocking algorithms assume the occupancy-bound
execution model, which they exploit by launching no more workgroups
than there are available compute units~\cite{owens-persistent}. This
works because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

@@

@@

\item Cooperative kernels can be implemented over the occupancy\-/bound
  execution model that current GPUs provide: our prototype implementation requires
  no additional hardware or driver support.  Thus the model should be easy for vendors to implement directly.

@@

>>>>

\newcommand{\myparagraph}[1]{\paragraph{#1.}}
%{\noindent{\bf\emph{#1}}\hspace{1em}}

@@

\myparagraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating
such algorithms on GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups.  For example, many graph algorithms employ a level-by-level
strategy, requiring a global barrier between each level, and work
stealing algorithms require each workgroup to maintain a queue, which
is typically protected by a mutex to enable stealing by other
workgroups.

To avoid starvation, a blocking concurrent algorithm requires
\emph{fair} scheduling of workgroups.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.

%\vspace{-1mm}
\myparagraph{A degree of fairness: occupancy-bound execution} The current GPU programming
models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding
scheduling of workgroups, and current GPU schedulers are unfair in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.\footnote{In practice,
  depending on the kernel, multiple workgroups might map to the same
  compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle more workgroups being
launched than there are compute units is via an \emph{occupancy-bound}
execution
model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where,
once a workgroup has commenced execution on a compute unit (it has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it finishes execution.  
This is simple model is easy to implement
efficiently, and experiments suggest that
it is widely employed by today's
GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{overview.pdf}
\caption{Cooperative kernels can flexibly be resized in order to let
other tasks (e.g. graphics) access GPU ressources}
\label{fig:overview}
\end{figure}

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, which are
bound to separate compute units that operate in parallel.  Current GPU
implementations of blocking algorithms assume the occupancy-bound
execution model, which they exploit by launching no more workgroups
than there are available compute units~\cite{owens-persistent}.

@@

For example, Nvidia's latest architecture (Pascal) is documented to
support instruction-level preemption~\cite{??}. Despite this, Nvidia's
most recent compute capablity does not guarantee forward progress for
CUDA programmers (evidence that vendors will not yet commit to providing such guarantees).
Indeed, on a GTX Titan X (Pascal) we still observe starvation:
we found that a global barrier can be executed successfully with 56 workgroups, but
deadlocks with 57 workgroups.  This is an example of a vendor
supporting preemption, yet not guaranteeing fair scheduling.

@@

\item Cooperative kernels can be implemented over the occupancy\-/bound
  execution model that current GPUs provide: our prototype implementation requires
  no additional hardware or driver support.

\item If hardware support for preemption is available it can be leveraged to implement cooperative kernels efficiently, and cooperative kenrnels can avoid the overhead of unnecessary preemptions by allowing the programmer to communicate ``smart'' preemption points.

@@

\subsection{Comparison with Task Level Preemption}\label{sec:nvidiacomparison}

Some recent GPUs, including NVIDIA's Pascal architecture, provide
hardware support for instruction-level preemption.  Few details of how
this works are available, however from reading NVIDIA documentation
and online material~\cite{ }, and running litmus tests on a Pascal
platform, our understanding is that preemption of entire kernels, but
not of individual workgroups running a kenrel is supported.
Intel---the only vendor whose OpenCL implemenation is capable enough
to implement cooperative kernels---do not provide this feature, thus
we cannot make a direct comparison.  We present here a theoretical
analysis of the overheads associated with sharing the GPU between
graphics and compute tasks via kernel-level preemption.

Let us assume that a graphics workload is required to be scheduled
with period $P$ and duration $D$, and that a long-running compute
kernel requires time $C$ to execute without interruption.  If we
assume the cost of preemption is negligible (e.g.\ NVIDIA have
reported preemption times of 0.1 ms for Pascal), then the overhead
associated with switching between compute and graphics every $P$ time
steps is $P/(P-D)$.

\ADComment{Tyler to complete.}
